Effects of pre-existing structures on the seismicity of the Charlevoix Seismic Zone

### 2019-06-28 Rebuttal_Letter.tex

The following are our detailed responses to the comments by the reviewers:

\Reviewer{\#1}

\begin{response}{The manuscript of Fadugba et al. "Effects of pre-existing structures on the seismicity of the Charlevoix Seismic Zone" introduces a modeling study assessing the stress perturbation due to a specific tectonic structure, that is, a rift system overprinted on the impact crater in Charlevoix Seismic Zone, eastern Canada. The effects of possible factors, including fault geometry, frictional strengths along faults, elastic properties of the crater and the different velocity models, are being discussed by comparing the causal stress perturbations to relocated seismic distributions. They study the combined effects of the impact zone and rift faults, confirm the importance of the two tectonic factors in explaining the stress perturbation, and provide a quantified estimation of the elastic contrast within and outside the impact crater. Overall, I find that their research approach is well designed. While the result is clearly presented, my main concern is how to justify conclusions based on model comparisons with single factor control.}
\end{response}

\begin{response}{I recommend the authors to justify the assumption of the independency effect of involved factors on stress perturbations, or that they have consistently monotonous effects.}
    If the term, ``the independency effect of involved factors'', means the effects of independently controlled parameters on model results. The independency is never assumed and only a logical outcome from our experiment design, in which a reference model is compared with a model with only one parameter changed.

@@

As for the monotonicity of model responses to varied parameters, our original results are already showing that that the controlled parameters have consistently monotonous effects. However, in this revision, we have run more models so that all the explored parameters have more regular intervals and thus the monotonicity of model responses can be exposed more clearly. For more details about the changes we have made in this regard, please refer to our response right below as well as those to the reviewer's specific comments.
\end{response}

\begin{response}{Also, I would be curious to know the sensitivity of your model to these factors, so that we will have a rough idea of the model resolution. It will also support your choice of parameter intervals, for example, moduli ratio and $\mu$ (Table 1). Thus, I suggest a moderate revision.}

    We have addressed this concern by running new models, making parameter intervals more regular. The expanded set of models not only provide the sensitivity information but also confirm that the model responses are monotonous to the control parameters in the considered ranges of values. Furthermore, the revised manuscript includes an updated justification for the choice of model parameters such as friction coefficient and elastic moduli ratio (Lines 368-374 and 422-425) as well as  in the supplementary information (Figs. S3-6) 
    %\note[EC]{Seun, specify where in the revised manuscript to look at for this information.} %\note[EC]{Why S5 is not listed?}
\end{response}

@@

We have added Fig. S1 in the supplementary information.

\end{response}

\begin{response}{2. Line 116: Introduce the computational efficiency of your model, like how much cores your model requests, how long it takes. You can add this information into Table 1 if it varies much between different models.}

  We have added the computational efficiency of our model in the model geometry section (Lines 131-137) of the manuscript. We did not add the computational efficiency to Table 1 because it does not change significantly among the models except for the SD65R25 model, which took about twice longer to run for the same model time. We do not clearly understand why this model is particularly slow but further analyzing computational efficiency is beyond the scope of this paper.

@@

  %\add[EC]{, which took about twice longer to run for the same model time.} \annote[EC]{which may be due to the shallower fault dips.}{The shallower fault dips are the difference of SD65R25 from the others but does not directly explain why this model should take much less time. We can go ahead and speculate about possible reasons but it would be just enough to say that we do not clearly understand why this model is particularly slow and further analyzing computational efficiency is beyond the scope of this paper.}
  %\annote[EC]{except for the SD65R25 model}{This model then needs further explanation why it took a significantly longer or short time}. 
  %\note[EC]{Is there a reason why you did not follow the reviewer's suggestion to add this into to Table 1?}
\end{response}

\begin{response}{3. Line 131-133: I do not understand why you set your boundary conditions as the differential stress reaches 706 MPa. Do you set a criterion for shear stress somewhere else? Please make it more clear.}

  %\annote[EC]{We agree with the reviewer}{Agree only when you are sure that doing so would not negatively affect your arguments, assumptions or the overall integrity of your paper. Never confess; only justify.} on the high differential stress at seismogenic depth in our study, due to}
  The differential stress of 706 MPa at 10 km depth is based on Byerlee's law under the assumption of  
  %$\mu$ equal to 0.6 and \add[OF]{The equation of Byerlee's law has a μ value of 0.6. It is not an assumption in our study.}
  %\add[EC]{$\mu$ equal to 0.6 and} 
  the absence of pore-fluid pressure. The full derivation of this value is now provided in the supplementary information. Shear stress was not considered as a stopping criterion. In addition, we have included a discussion on the possibility and some consequences of pore-fluid pressure at depth (Lines 534-546).

@@

\begin{response}{4. Section 2.3: You did not mention the values of bulk modulus and shear modulus applied in your calculation.}

  PyLith, the modeling software used in this study, internally converts P and S wave velocities and density to bulk and shear moduli. For readers' reference, however, we now list representative values of elastic moduli in Table 1. 
  %\add[EC]{For readers' reference, however, we now list representative values of elastic moduli in Table 1.}\note[EC]{Seun, please take care of this.}
\end{response}

\begin{response}{5. Line 154-156: As I said in the main concern, you need to justify the reduction of 10\% from the velocity model to represent the impact crater area, instead of, for example, 2\%, 5\%, or 20\%.}

  The revised manuscript now includes a justification that the 10\% reduction of the density in the impact structure represents damage made to crustal rocks by the impact (Lines 181-186). Since it is not known quantitatively how much the impact affected the crustal density and elastic moduli, we ran extra models to explore the effects of density reduction from 0 to 20 \%. No significant changes in the model results were found in these models. These new results are now included in the supplementary information and Fig. S3 therein.
\end{response}

@@

    We have explored non-zero values of cohesion and their impact on the results are shown in Fig. S4 in the supplementary information. However, since low-cohesion results show a better spatial correlation with seismicity, the revised manuscript still presents models with zero cohesion as the main results.

\end{response}

\begin{response}{7. Line 226-231: When evaluate the modeling result, I suggest the authors to quantify the fitting between the simulated stress change and the earthquake hypocenters. It will help convincing the readers of preferred model parameters. For example, I do not see it necessarily to prefer the frictional coefficient equals to 0.3 over 0.4 based on Figure 11.}

   We have added a quantitative assessment of the correlation between the change in differential stress and the observed seismicity by determining the percentage of hypocenters in the region with $\Delta\sigma_{D}^\prime \ge$ 0.5\% without accounting for uncertainty in hypocenter relocation (Line 205-209). This metric is now used for describing the model results and comparing different models throughout the revised manuscript. We also added a discussion on the choice of the friction coefficient in Lines 512-518.
    %: 1) the change in differential stress maps changes with fault geometry, and we have established an along-strike variation in the fault dips in this study; and 2) the 3D velocity model is very well defined in the upper 5 km depth which may introduced an uncertainty in the hypocenter relocation.
\end{response}

@@

\begin{response}{8. Line 312: You need to justify the mesh size of 4 km in Section 2.1.}

We decided to run this particular group of models to acquire necessary information without increasing computational cost too much. Low-friction models, especially those with $\mu$ equal to 0.1 and 0.2, are very expensive 
%take \annote[EC]{up to XX weeks}{Seun, please fill out this part} 
due to slow convergence rate in the non-linear solution scheme. Running all the models in this group at the same mesh resolution of 4 km ensures intra-group consistency; and comparison with 2 km-resolution models is also justified because the descritization error does not dramatically increase as the resolution increases to 4 km (Fig. S1). This discussion is now included in Lines 118 to 127 of the revised manuscript. % \annote[EC]{XXX-YYY}{Seun, please do add this discussion and fill out the line numbers here} 
\end{response}

@@

\begin{response}{An aspect that may have been missed is the fact that the epicentral map, such as Fig. 1, does not show a complete view of the regional seismicity. The SSW portion of the St. Lawrence River (upstream towards Quebec City) is also active (although not as much as the interior of the astrobleme). The dataset shown is the one of Powell and Lamontagne (2017) which was a selection of events based on their locations within the network of seismograph stations, but for this map view, I suggest that a more complete dataset is presented. The data can be downloaded from the earthquakescanada web site.}

  We have plotted the complete dataset in Fig. 1. We represented the relocated hypocenters of Powell and Lamontagne (2017) as circles and the other earthquakes that are not in the relocated catalog as diamonds.
\end{response}

\begin{response}{In a couple of places, authors use the expression 'large magnitude earthquakes'... Except for the 1925 M 6.2 earthquake, we are uncertain about the exact position of the 'large' earthquakes in Charlevoix (four other M6+; 1663; 1791, 1860,1870... although 1870 M 6.5 was probably near Ile-aux-Coudres). It would be better to say that the M4+ earthquakes, appear to occur more often in the NE portion of the CSZ. Stevens (1980) also found that some M4+ earthquakes of the period 1924-1980 also occurred at the SE end.}

  We have edited the manuscript accordingly (see Lines 454-456). We have also included the findings of Steven (1980) in our discussion.
\end{response}

@@

\begin{response}{The authors write that the focal mechanisms of the CSZ were not well defined. I disagree with them. You can check this OF report, which shows that the Quality 'A' mechanisms were well defined by the P first motions. I think that Mazzotti used most of them in his catalog.
https://geoscan.nrcan.gc.ca/starweb/geoscan/servlet.starweb?
path=geoscan/downloade.web\&search1=R=209378}

 The focal mechanisms are good, and we suggest that the focal mechanisms can be further constrained using waveform modeling to validate the magnitude of the stress rotation (Lines 569-572).
\end{response}

@@

\begin{response}{The model assumes an absence of pore-fluid pressure along faults. For this reason, the stress difference used by the authors is very high at seismogenic depth. It is my impression that worldwide, direct measurements and modelling of tectonic stresses suggest that the maximum crustal stress difference has an upper limit of about 100-200 MPa... very different from the 706 MPa suggested by the authors.
I think that the authors should discuss the possibility that high pore-fluid pressures exist at depth and their possible consequences. One aspect would be that faults could be reactivated at smaller angles from the maximum compressive stress and that the stress difference would have more realistic values.}

  The differential stress of 706MPa at 10 km depth is due to the assumption of the absence of pore-fluid pressure. The higher stopping criterion in our study is necessary to make faults slip at seismogenic depth under the assumption. We have included a discussion on the possibility and some consequences of pore-fluid pressure at depth (Lines 534-546).
\end{response}

### 2019-06-28 main_full_manuscript.tex
Addressed remaining comments in the Rebuttal letter and have updated the manuscript

We conducted resolution tests in which the change in displacement and stress solutions relative to a 1 km-resolution model are monitored as a function of element size. \change[OF]{Displacement was almost constant for 2 km or smaller element sizes}{Displacements and differential stresses do not change appreciably for elements sizes equal to 4 km or smaller (Fig. S1)}. \add[OF]{Based on the resolution tests, we use the element size of 2 km for most of the models and 4 km for more computationally expensive models like those for exploring the effects of the friction coefficient.} We also observe edge effects by comparing a 420 $\times$ 420 $\times$ 40 km model with the original 220 $\times$ 220 $\times$ 40 km model, finding that the seismogenic parts around the \change[OF]{crater}{impact structure} show consistent displacement and stress solutions for both domain sizes. Based on this result, we choose the smaller domain for this study and set the element size to be about 2 km. \add[OF]{The average computation time was about 16 days for fault models ($\mu$ = 0.3) with 2 km element size on 16 cores on the University of Memphis High Performance Computing cluster. The model with fault dips of 65$^\circ$-40$^\circ$-40$^\circ$ took about two times as long to run for the same model time. We do not clearly understand why the model with shallower-dipping faults is slower than the others roughly by a factor of two and further analyzing computational efficiency is beyond the scope of this paper.}
%we use an element size of 4 km for models with friction coefficient ($\mu$) values of 0.1 and 0.2 due to their expensive computation time.

@@

\subsection{Earthquake distribution}
Our models suggest that having both the rift faults and the impact \change[OF]{crater}{structure} is not only sufficient but also necessary for explaining the distribution of earthquakes in the CSZ. The models without a weaker \change[OF]{crater}{impact structure} zone, SD70R100 and TD70, show regions of positive $\Delta \sigma_{D}^{\prime}$, a proxy for earthquake generation potential, along the entire length of the rift faults (e.g., Fig. \ref{fig:eff_of_moduli_ratio}B). In contrast, when the \change[OF]{crater}{impact structure} is set to be \change[OF]{less elastically stiff}{elastically softer} than the surrounding crust, positive $\Delta \sigma_{D}^{\prime}$ regions spatially coincide with the observed seismicity in the CSZ, especially above 15 km depth (e.g., Fig. \ref{fig:SD70R25}C). Our models predict that earthquake activity should extend to about 20 km \add[OF]{distance} from the \change[OF]{crater}{impact structure} in a northeast direction (Fig. \ref{fig:SD70R25}C). This distribution of positive $\Delta \sigma_{D}^{\prime}$ is not seen in the reference model without the rift faults. These two findings from our models are consistent with the conclusion drawn by~\citet{Thomas_Powell_2017} that the seismicity in CSZ requires both the damaged \change[OF]{crater}{impact structure} zone and the rift faults.

>>>>

We conducted resolution tests in which the change in displacement and stress solutions relative to a 1 km-resolution model are monitored as a function of element size. \change[OF]{Displacement was almost constant for 2 km or smaller element sizes}{Displacements and differential stresses do not change appreciably for elements sizes equal to 4 km or smaller (Fig. S1)}. \add[OF]{Based on the resolution tests, we use the element size of 2 km for most of the models and 4 km for more computationally expensive models like those for exploring the effects of the friction coefficient. Low-friction models, especially those with $\mu$ equal to 0.1 and 0.2, are very expensive due to slow convergence rate in the non-linear solution scheme. Running all the models in this group at the same mesh resolution of 4 km ensures intra-group consistency; and the comparison with 2 km-resolution models is also justified because the descritization error does not dramatically increase as the resolution increases to 4 km (Fig. S1).} We also observe edge effects by comparing a 420 $\times$ 420 $\times$ 40 km model with the original 220 $\times$ 220 $\times$ 40 km model, finding that the seismogenic parts around the \change[OF]{crater}{impact structure} show consistent displacement and stress solutions for both domain sizes. Based on this result, we choose the smaller domain for this study and set the element size to be about 2 km. \add[OF]{The average computation time was about 16 days for fault models ($\mu$ = 0.3) with 2 km element size on 16 cores on the University of Memphis High Performance Computing cluster. The model with fault dips of 65$^\circ$-40$^\circ$-40$^\circ$ took about two times as long to run for the same model time. We do not clearly understand why the model with shallower-dipping faults is slower than the others roughly by a factor of two and further analyzing computational efficiency is beyond the scope of this paper.}
%we use an element size of 4 km for models with friction coefficient ($\mu$) values of 0.1 and 0.2 due to their expensive computation time.

@@

\subsection{Earthquake distribution}
Our models suggest that having both the rift faults and the impact \change[OF]{crater}{structure} is not only sufficient but also necessary for explaining the distribution of earthquakes in the CSZ. The models without a weaker \change[OF]{crater}{impact structure} zone, SD70R100 and TD70, show regions of positive $\Delta \sigma_{D}^{\prime}$, a proxy for earthquake generation potential, along the entire length of the rift faults (e.g., Fig. \ref{fig:eff_of_moduli_ratio}B). In contrast, when the \change[OF]{crater}{impact structure} is set to be \change[OF]{less elastically stiff}{elastically softer} than the surrounding crust \add[OF]{ by up to a factor of 4}, positive $\Delta \sigma_{D}^{\prime}$ regions spatially coincide with the observed seismicity in the CSZ, especially above 15 km depth (e.g., Fig. \ref{fig:SD70R25}C). Our models predict that earthquake activity should extend to about 20 km \add[OF]{distance} from the \change[OF]{crater}{impact structure} in a northeast direction (Fig. \ref{fig:SD70R25}C). This distribution of positive $\Delta \sigma_{D}^{\prime}$ is not seen in the reference model without the rift faults. These two findings from our models are consistent with the conclusion drawn by~\citet{Thomas_Powell_2017} that the seismicity in CSZ requires both the damaged \change[OF]{crater}{impact structure} zone and the rift faults.
